---
title: "More Data"
author: "Nidhi Ram"
date: "2025-06-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyverse)
library(sf) 
library(ggplot2) 
library(leaflet) 
library(dplyr) 
library(tigris) 
library(tidyr)
library(gt)
library(zipcodeR)
library(maps)
library(tidycensus)
library(terra)
library(exactextractr)
library(raster)
library(tidycensus)
library(ggthemes)
library(units)
```

```{r, userdirectory, echo = TRUE, message=FALSE, warning=FALSE}

# Set user directory and folder
user <- "/Users/nidhiram"
directory <- "/Downloads"
folder <- "/Summer 2025 Project" 

# Set working directory
working_directory <- "/Users/nidhiram/Downloads/Summer 2025 Project"
print(working_directory)
setwd(working_directory)

```

# Population Density
```{r}
population_count <- rast("data/2020_Dasymetric_Population_CONUS_V1_1/2020_Dasymetric_Population_CONUS_V1_1.tif")
us_grid2 <- st_transform(us_grid, st_crs(population_count))
us_grid2$pop_sum <- exact_extract(population_count, us_grid2, "sum")
us_grid2 <- st_transform(us_grid2, 5070)


us_grid$population <- us_grid2$pop_sum
us_grid$pop_density <- us_grid$population / us_grid$area_km2


ggplot(us_grid) +
  geom_sf(aes(fill = pop_density)) +
  scale_fill_viridis_c(option = "C") +
  theme_minimal() +
  labs(title = "Population Density",
       fill = "Density")

minnesota_counties$pop_sum <- exact_extract(population_count, minnesota_counties, "sum")
minnesota_counties <- minnesota_counties %>%
  mutate(area_m2 = st_area(geometry), area_km2 = area_m2 / 1e6, pop_density = pop_sum / area_km2)
minnesota_counties$pop_density <- as.numeric(minnesota_counties$pop_density)

ggplot(minnesota_counties) +
  geom_sf(aes(fill = pop_density)) +
  scale_fill_viridis_c(option = "C") +
  theme_minimal() +
  labs(title = "Population Count (MN)",
       fill = "People per County")

st_write(mn_counties2, "minnesota.shp")

```

# Environmental Justice Index
```{r}
eji <- read.csv("data/EJI_2024_United_States_CSV/EJI_2024_United_States.csv")
eji$COUNTYFP  <- sprintf("%03s", eji$COUNTYFP)
eji$STATEFP <- sprintf("%02s", eji$STATEFP)

eji <- eji %>% 
  mutate(pop_eji = E_TOTPOP * RPL_EJI, FIPS = paste0(STATEFP, COUNTYFP))


county_eji <- eji %>%
  group_by(FIPS) %>%
  summarise(mean_EJI = mean(RPL_EJI, na.rm=TRUE),
    pop_weighted_EJI = sum(pop_eji, na.rm=TRUE) / sum(E_TOTPOP, na.rm=TRUE))

minnesota_counties <- left_join(minnesota_counties, county_eji, by="FIPS")
minnesota_grid <- left_join(minnesota_grid, county_eji, by="FIPS")
minnesota_grid <- subset(minnesota_grid, select = -c(pop_weighted_EJI))


us_counties_midwest <- left_join(us_counties_midwest, county_eji, by="FIPS")
midwest_grid <- left_join(midwest_grid, county_eji, by="FIPS")
midwest_grid <- subset(midwest_grid, select = -c(pop_weighted_EJI))

us_counties_contiguous <- left_join(us_counties_contiguous, county_eji, by="FIPS")
us_grid <- left_join(us_grid, county_eji, by="FIPS")
us_grid <- subset(us_grid, select = -c(pop_weighted_EJI))

```


# Agricultural Employment and Wages
```{r}
library(tidycensus)
library(tidyverse)
library(viridis)
tract_employment <- get_acs(geography = "tract", variables = c("C24030_001", "C24030_032", "C24030_005"), year = 2023, output= "wide", state = contiguous_states, geoemtry=TRUE)
tracts <- tracts(cb = TRUE, refresh=TRUE)
tracts <- st_transform(tracts, crs = 5070)
tracts$STATEFP  <- sprintf("%02s", tracts$STATEFP)
tracts$COUNTYFP <- sprintf("%03s", tracts$COUNTYFP)
tracts$FIPS <- paste0(tracts$STATEFP, tracts$COUNTYFP)

tract_employment <- left_join(tract_employment, tracts, by = "GEOID")
tract_employment <- subset(tract_employment, select = -c(STATEFP, COUNTYFP, TRACTCE, GEOIDFQ, NAMELSAD, ALAND, AWATER))
tract_employment <- subset(tract_employment, select = -c(NAME.y, LSAD, NAME.x))

tract_employment <- subset(tract_employment, select = -c(C24030_001M, C24030_032M, C24030_005M))
tract_employment <- tract_employment %>% rename(total_employment = C24030_001E, male_ag_employment = C24030_005E, female_ag_employment = C24030_032E) %>%
  rename(COUNTY = NAMELSADCO)
tract_employment <- tract_employment %>%
  mutate(ag_employment = male_ag_employment + female_ag_employment) %>%
  mutate(pct_ag = 100 * ag_employment / total_employment) 
tract_employment$pct_ag <- tract_employment$pct_ag / 100
tract_employment <- subset(tract_employment, select = -c(geometry))



```

## Assign to Grid
```{r}
us_grid_centroids <- st_centroid(us_grid)
grid_with_geoid <- st_join(us_grid_centroids, dplyr::select(tracts, GEOID), join = st_within) 
missing <- is.na(grid_with_geoid$GEOID)
grid_intersect <- st_join(us_grid[missing, ], dplyr::select(tracts, GEOID), join = st_intersects, left=FALSE)
grid_with_geoid$GEOID <- ifelse(missing, grid_intersect$GEOID, grid_with_geoid$GEOID)
us_grid$GEOID <- grid_with_geoid$GEOID

us_grid <- left_join(us_grid, tract_employment, by = "GEOID")
us_grid <- subset(us_grid, select = -c(FIPS.y))
us_grid <- st_transform(us_grid, crs = 5070)
us_grid <- us_grid %>%
  dplyr::select(cropland:uninhabited, Number_of_Farms:pct_ag, FIPS = FIPS.x, geometry)

```


## distance to cluster
```{r}
tract_employment <- left_join(us_grid, tract_employment, by = "GEOID")

ag_centroids <- us_grid %>%
  mutate(ag_employment = replace_na(ag_employment, 0)) %>%
  st_centroid()
threshold <- quantile(ag_centroids$ag_employment, 0.9, na.rm = TRUE)
ag_clusters <- ag_centroids %>% filter(ag_employment >= threshold)

dist_matrix <- st_distance(st_centroid(us_grid), st_centroid(ag_clusters))

# Take minimum distance for each grid cell
us_grid$dist_to_ag_cluster_km <- apply(dist_matrix, 1, min) / 1000  # in km

```

# Midwest TRI
```{r}
mw_toxic_releases <- read.csv("midwest_tri.csv")
mw_toxic_releases <- st_as_sf(mw_toxic_releases, coords = c("Longitude", "Latitude"),crs = 5070)

mw_toxic_releases <- st_transform(mw_toxic_releases, 5070)
mw_toxic_releases <- mw_toxic_releases %>% rename(`releases` = `Releases..lb.`)
mw_toxic_releases$releases <- decomma(mw_toxic_releases$releases)

mw_toxic_releases$releases <- as.numeric(mw_toxic_releases$releases)
summary(mw_toxic_releases$releases)
hist(mw_toxic_releases$releases)

mw_toxic_releases$releases[is.na(mw_toxic_releases$releases)] <- 0
us_counties_midwest |> 
  st_union() -> Midwest

ggplot() +
  geom_sf(
    data = Midwest,
    fill = "white",
    color = "black"
    ) +
  geom_sf(
    data = mw_toxic_releases,
    aes(color = releases))


library(gstat)
idw(formula = releases ~ 1, locations = mw_toxic_releases, newdata = midwest_grid, idp = 2.5) -> idw_output
summary(idw_output$var1.pred)
hist(idw_output$var1.pred)
library(classInt)
nbreaks <- classIntervals(idw_output$var1.pred, n = 5, style = "fisher")$brks

ggplot() +
  geom_sf(
    data = Midwest,
    fill = "white",
    color = "black"
  ) +
  geom_sf(
    data = idw_output,
    aes(fill = var1.pred),
    color = NA,
    alpha = .9
  ) +
  scale_fill_fermenter(
    direction = 1,
    palette = "YlOrRd",
    breaks = nbreaks
  )

midwest_grid <- st_join(midwest_grid, idw_output, left = TRUE,largest = TRUE)
midwest_grid <- subset(midwest_grid, select = -c(x, y, var1.var))

bg_employment <- get_acs(geography = "block group", variables = c("C24030_001", "C24030_032", "C24030_005"), year = 2023, output= "wide", state = c("17", "18", "19", "20", "26", "27", "29", "31", "38", "39", "46", "55"), geoemtry=TRUE)

```

# states with cafos
Maryland, Pennsylvania, Iowa, Missouri, Minnesota, Indiana, Idaho, Washington, Ohio, Michigan, Texas, North Carolina, South Carolina, Virginia

NEED: Oregon, California, Delaware, Illinois, Nebraska, Nevada

Downloaded: LA, NY, AL, AR, CO, FL, GA, MS, WI
```{r}

cafo_states <- c("Maryland", "Pennsylvania", "Iowa", "Missouri", "Minnesota", "Indiana", "Idaho", "Washington", "Ohio", "Michigan", "Texas", "North Carolina", "South Carolina", "Virginia", "Oregon", "California", "Delaware", "Illinois", "Nebraska", "Nevada", "Louisiana", "New York", "Alabama", "Arkansas", "Colorado", "Florida", "Georgia", "Mississippi", "Wisconsin")

cafo_states <- us_counties_contiguous %>%
    filter(STATE_NAME %in% cafo_states)

ggplot() +
  geom_sf(data = us_counties_contiguous, fill = "lightblue", color = "black", size = 0.2) +
  geom_sf(data = cafo_states, aes(fill = "CAFOs")) + 
  scale_color_manual(values = c("CAFOs" = "red")) +
  labs(title = "Publicly Available CAFO Locations in CONUS",
       fill = "Legend") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 
ggsave(filename = "CAFO_states.jpg", width = 8, height = 5)


```


# canpoy cover
```{r}
canopy_cover <- rast("data/nlcd_tcc_CONUS_2015_v2023-5_wgs84/nlcd_tcc_conus_wgs84_v2023-5_20150101_20151231.tif")
us_grid$canopy_cover <- exact_extract(canopy_cover, us_grid, "sum")
us_grid <- st_as_sf(us_grid)
```


# urban area proximity
```{r}
urban_areas <- st_read("data/NTAD_Urban_Areas_644170322766021823/Urban_Areas.shp")
urban_areas <- st_transform(urban_areas, 5070)
dist_matrix <- st_distance(st_centroid(us_grid), st_centroid(urban_areas))

us_grid$dist_to_urban <- apply(dist_matrix, 1, min) / 1000  # in km


```


# meat, poultry, egg facilities
```{r}
meat_facilities <- read.csv("data/mpe_facilities.csv")
meat_facilities <- meat_facilities %>%
  filter(!is.na(longitude)) %>%
  filter(!is.na(latitude))

meat_facilities <- st_as_sf(meat_facilities, coords = c("longitude", "latitude"), crs = 5070)
dist_matrix <- st_distance(st_centroid(us_grid), meat_facilities)
us_grid$dist_to_facility <- apply(dist_matrix, 1, min) / 1000  # in km


```


# Soil cores
```{r}
library(usmap)
soil_cores <- read_xls("data/Appendix_2a_Top5_18Sept2013.xls", col_names=TRUE, skip=12)
soil_cores <- soil_cores[-1, ]
soil_cores <- st_as_sf(soil_cores, coords = c("Longitude", "Latitude"), crs = 5070)

phosphorus <- subset(soil_cores, select = c(StateID, Top5_P, geometry))
phosphorus$state_fips <- fips(state = phosphorus$StateID)

phosphorus <- phosphorus %>% filter(state_fips %in% contiguous_states)
phosphorus$Top5_P <- as.numeric(phosphorus$Top5_P)
st_write(phosphorus, "phosphorus.shp", row.names = FALSE)
phosphorus$Top5_P[is.na(phosphorus$Top5_P)] <- 0

# idw
idw(formula = Top5_P ~ 1, locations = phosphorus, newdata = us_grid, idp = 3, debug.level = 1, nmax = 15) -> idw_output
summary(idw_output$var1.pred)
hist(idw_output$var1.pred)
library(classInt)
nbreaks <- classIntervals(idw_output$var1.pred, n = 5, style = "fisher")$brks

ggplot() +
  geom_sf(
    data = US,
    fill = "white",
    color = "black"
  ) +
  geom_sf(
    data = idw_output,
    aes(fill = var1.pred),
    color = NA,
    alpha = .9
  )
us_grid$phosphorus <- idw_output$var1.pred

```


# census data
```{r}
census_socio <- get_acs(geography = "county", variables = c("B01001B_001"),
  year = 2023,
  survey = "acs5")
census_socio2 <- get_acs(geography = "county", variables = c("B01001I_001"),
  year = 2023,
  survey = "acs5")
census_socio$Black_Hispanic <- census_socio$estimate + census_socio2$estimate

census_socio <- subset(census_socio, select = -c(moe, variable, NAME))
census_socio2 <- subset(census_socio2, select = -c(moe, variable, NAME))
census_socio <- census_socio %>% rename(Black = estimate)
census_socio2 <- census_socio2 %>% rename(Hispanic = estimate)

midwest_grid <- left_join(midwest_grid, census_socio, by = c("FIPS" = "GEOID"))
midwest_grid <- left_join(midwest_grid, census_socio2, by = c("FIPS" = "GEOID"))
us_grid <- left_join(us_grid, census_socio, by = c("FIPS" = "GEOID"))
us_grid <- left_join(us_grid, census_socio2, by = c("FIPS" = "GEOID"))


census_vars <- load_variables(2023, "acs5")

```


# manure nutrients
```{r}
manure <- read_excel("data/N-P_from_manure_1950-2017-july23-2020.xlsx", sheet = "2017")
manure <- subset(manure, select = -c(...25, `cattle-weight-coef`, `hogs-weight-coef`, `chickens-weight-coef`, `broilers-weight-coef`, `turkeys-weight-coef`, `sheep-weight-coef`, `fips-int`, CountyName, State))
midwest_grid <- left_join(midwest_grid, manure, by = c("FIPS" = "STCOFIPS"))
midwest_grid$animal_P <- midwest_grid$`Cattle_P_kg-2017` + midwest_grid$`Hogs_P_kg-2017` + midwest_grid$`Poultry_P_kg-2017`
midwest_grid$animal_N <- midwest_grid$`Cattle_N_kg-2017` + midwest_grid$`Hogs_N_kg-2017` + midwest_grid$`Poultry_N_kg-2017`
midwest_grid$animal_count <- midwest_grid$cows2017adj + midwest_grid$bcows2017adj + midwest_grid$mcows2017adj + midwest_grid$othercows2017 + midwest_grid$hogs2017adj + midwest_grid$chickens2017adj + midwest_grid$broilers2017adj + midwest_grid$turkeys2017adj + midwest_grid$sheep2017adj + midwest_grid$horses2017adj 


```

# air monitors
```{r}
air_monitors <- read.csv("data/annual_conc_by_monitor_2024.csv")
pm25 <- read.csv("data/daily_88101_2024.csv")
pm25 <- st_as_sf(pm25, coords = c("Longitude", "Latitude"), crs = 5070)
pm25 <- subset(pm25, select = -c(Parameter.Code, Parameter.Name, POC, Datum, Sample.Duration, Pollutant.Standard, Date.Local, Units.of.Measure, Observation.Count, Observation.Percent, Method.Code, Method.Name, Local.Site.Name, Address, State.Name, County.Name, City.Name, CBSA.Name, Date.of.Last.Change))
avg_pm25 <- pm25 %>% group_by(Site.Num) %>% summarise(avg_pm25 = mean(Arithmetic.Mean))


idw(formula = avg_pm25 ~ 1, locations = avg_pm25, newdata = midwest_grid, idp = 2) -> pm25_idw
summary(pm25_idw$var1.pred)
hist(idw_output$var1.pred)
library(classInt)
nbreaks <- classIntervals(idw_output$var1.pred, n = 5, style = "fisher")$brks


```

# assign fips 
```{r}
us_grid_centroids <- st_centroid(us_grid)

us_grid_with_fips <- st_join(us_grid_centroids,
  dplyr::select(us_counties_contiguous, FIPS),
  join = st_within)

# Step 2: Identify grid cells with no county assigned
us_missing <- is.na(us_grid_with_fips$FIPS)

# Step 3: Assign missing grid cells by spatial intersection using the **original grid geometry**
us_grid_intersect <- st_join(us_grid[us_missing, ],  
  dplyr::select(us_counties_contiguous, FIPS),
  join = st_intersects,
  left = FALSE)

# Step 4: Update FIPS values for missing grid cells
us_grid_with_fips$FIPS[us_missing] <- us_grid_intersect$FIPS

# Step 5: Transfer FIPS codes to the original grid
us_grid$FIPS <- us_grid_with_fips$FIPS


us_grid <- us_grid %>%
  left_join(us_counties_contiguous %>%
      dplyr::select(FIPS, Number_of_Farms) %>%
      st_drop_geometry(), by = "FIPS")

```

# save grid with features as csv
```{r}
write.csv(us_grid, "us_grid.csv")
st_write(us_grid, "us_grid.shp", row.names = FALSE)


```

# open saved grid
```{r}
us_grid <- st_read("us_grid/us_grid.shp")

names(us_grid)
us_grid <- us_grid %>% rename(cropland = croplnd, pastureland = pstrlnd, uninhabited = unnhbtd, RUCC_2023 = RUCC_20, population = popultn, canopy_cover = cnpy_cv, dist_to_ag_cluster_km = dst____, dist_to_facility = dst_t_f, phosphorus = phsphrs, Hispanic = Hispanc, animal_count = anml_cn, animal_N = animl_N, animal_P = animl_P, Other_P_kg_2017 = O_P_.20, Other_N_kg_2017 = O_N_.20, median_income = mdn_ncm, dist_to_urban = dst_t_r, female_ag_employment = fml_g_m, male_ag_employment = ml_g_mp)
us_grid <- us_grid %>% rename(mean_EJI = men_EJI)

```

# downscaling grid
```{r}
ggplot() +
  geom_sf(
    data = US,
    fill = "white",
    color = "black"
  ) +
  geom_sf(
    data = us_counties_contiguous,
  ) + theme_minimal()

ggplot() +
  geom_sf(
    data = US,
    fill = "white",
    color = "black"
  ) +
  geom_sf(
    data = us_grid,
  ) + theme_minimal()


```

# adjusting fips - state
```{r}
labeled$STATE_N[labeled$STATE_N == "Vermont"] <- "New York"
labeled$STUSPS[labeled$STUSPS == "VT"] <- "NY"
labeled$COUNTY[labeled$COUNTY == "Bennington County"] <- "Rensselaer County"

labeled$STATE_N[labeled$STATE_N == "Delaware"] <- "Texas"
labeled$STUSPS[labeled$STUSPS == "DE"] <- "TX"
labeled$COUNTY[labeled$COUNTY == "New Castle County"] <- "Chambers County"

labeled$COUNTY[labeled$COUNTY == "Hudspeth County"] <- "Hyde County"



```

# state level totals
```{r}
state_totals <- read.csv("data/state_totals_2024.csv", skip=6)
state_totals <- subset(state_totals, select = -c(X, X.1, X.2, X.3, X.4, X.5, X.6, X.7, X.8, X.9))
state_totals <- state_totals %>% rename(Total_CAFOs = Total.CAFOs1, CAFOs_with_NPDES_permits = CAFOs.with.NPDES.permits2)
state_totals <- head(state_totals, -5)

state_totals <- state_totals[state_totals$State != "Alaska", ]
state_totals <- state_totals[state_totals$State != "Hawaii", ]
state_totals <- state_totals[state_totals$State != "Indian Country (R7)", ]
state_totals <- state_totals[state_totals$State != "Indian Country (R8)", ]
state_totals <- state_totals[state_totals$State != "Puerto Rico", ]
state_totals <- state_totals[state_totals$State != "Virgin Islands", ]


state_totals_known <- labeled_df %>%
  group_by(STATE_N) %>%
  summarise(total_model = sum(cafo_count, na.rm = TRUE))
state_totals_known <- state_totals_known[state_totals_known$STATE_N != "Delaware", ]
state_totals_known <- state_totals_known[state_totals_known$STATE_N != "Maine", ]
state_totals_known <- state_totals_known[state_totals_known$STATE_N != "Tennessee", ]
state_totals_known <- state_totals_known[state_totals_known$STATE_N != "Massachusetts", ]


# join with totals --> THIS IS INTERESTING !!
state_totals_known <- state_totals_known %>%
  left_join(state_totals, by = c("STATE_N" = "State"))


```


